Modularly Composing Typed Language Fragments

### 2014-11-01 att-pldi15/att-pldi15.tex
More revisions through sec5

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%extensible languages; module systems; type abstraction; typed compilation; type-level computation
\section{Introduction}\label{intro}
Typed programming languages are most often described as being composed of \emph{fragments}, each contributing to the language's concrete syntax, abstract syntax, static semantics and dynamic semantics. 
In his textbook, Harper organizes fragments around type constructors, describing each in a different chapter \cite{pfpl}. Languages are then identified by a set of type constructors, e.g. $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language that includes partial function types, polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed below).
Other researchers introduce fragments by defining a simple calculus with a ``catch-all'' constant and base type to stand notionally for all other terms and types that may also be included in some future complete language.\todo{find a really good example, or add this to final PSP rev.??}

@@

This is not an everday problem for programmers only because fragments like those mentioned above are ``general purpose'': they make it possible to \emph{isomorphically embed}  many other fragments as ``libraries''. For example, list types need not be built in because they are isomorphic to the type $\iforall{\alpha}{\imu{t}{\iunit + (\alpha \times t))}}$ (e.g. datatypes in ML, which combine these into a single declaration construct). %Languages providing datatypes al\'a ML  are perhaps most directly oriented around  embeddings into this core. % Leaving the core language simple makes it easier to establish its metatheory and verify that tools, like compilers, are implemented correctly.% They also have encouraging connections to logic. %For example, we do not need to define the type constructor $\fvar{list}$ (indexed by a type) as a fragment (though it is possible to do so) because there is a user-defined datatype constructor, $\texttt{list}$, parameterized by a type, that is isomorphic. If we added the $\fvar{list}$ fragment to the language, it would be entirely redundant semantically: every well-typed term of a type constructed by $\fvar{list}$ corresponds to a well-typed term and a polymorphic recursive sum type such th%Although strings and numbers can be embedded as recursive sum types, it is recognized that this is impractical, so these are also usually included as primitives.
%Establishing an isomorphic embedding of a desirable fragment in terms of general-purpose fragments is not always possible, nor are such embeddings always practical.

Nevertheless, situations do arise where it is not possible to  use these fragments to establish an isomorphic embedding that preserves a desirable fragment's  static and dynamic semantics, including  performance bounds specified by a cost semantics. 
Embeddings are also sometimes too \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering various human factors. %We will discuss specific examples below.
%When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, 
%providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}. To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec. \ref{prior-work}. %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped. 
Each time a language is extended with a new fragment for one of these reasons, a new \emph{dialect} is born. Within the ML lineage, for example, dialects abound:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%

@@

This \emph{dialect-oriented} state of affairs is unsatisfying. %Language designers are burdened with needing to understand the complete metatheory of every fragment they add to their language to make sure it is not incompatible with existing fragments, so languages change rarely. This decreases the impact of many potentially useful innovations designed by fragment providers (e.g. many academic researchers) by making them unavailable to programmers. 
While programmers can choose from dialects supporting, e.g., a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, there may not be an available dialect supporting both. Using different dialects separately for different components of a program is untenable: components written in different dialects cannot always interface safely (i.e. a safe FFI, item 3 above, is needed).

These problems do not arise when a fragment is expressed as an isomorphic embedding (i.e. as a library) because modern \emph{module systems} enforce abstraction barriers that ensure that the isomorphism need only be established in the ``closed world'' of the module. There are no ``link-time'' proof obligations for clients in the ``open world''. For example, a module defining the semantics of sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g. uniqueness, if using a list representation) will hold no matter which other modules are separately in use by a client. %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec. \ref{prior-work} by summarizing them.

@@

\paragraph{Contributions} In this paper, we take foundational steps towards this goal by  constructing a minimal but powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus). %Despite its minimality, it can host a variety of practical semantic extensions like those described above, while maintaining strong metatheoretic guarantees and, crucially, providing modular reasoning properties. %a calculus introduced briefly in recent work on \emph{active type constructors}  \cite{atlang-gpce14}\todo{TR? Arxiv?}, reviewed in Sec. \ref{overview}. 
Its semantics are structured like those of many modern languages, consisting of an \emph{external language} (EL) governed by a {typed translation semantics} targeting a much simpler \emph{internal language} (IL). 
Rather than building in a monolithic set of external type constructors, however, the  semantics are indexed by a \emph{tycon context}. Each tycon defines the semantics of its operators via functions written in a \emph{static language} (SL) where types and translations are values. %, which provides a form of type-level computation. %The authors demonstrate the expressive power of this technique primarily with an implementation and a number of examples of fragments as libraries, and their core calculus .% (unlike systems that treat the type system as a ``bag of rules'', where non-determinism can arise).

We will begin by giving an overview of the organization  and main judgements of @$\lambda$ in Sec. \ref{atlam}, then discuss how types are constructed and introduce our two main examples, one defining labeled product types with a functional update operator, and the other regular string types, based on a recent core calculus style specification \cite{sanitation-psp14}, in Sec. \ref{types}. We describe how tycons define their associated term-level operator constructors (opcons) in Sec. \ref{external-terms}.  We next give the key metatheoretic properties of the calculus in Sec. \ref{metatheory}, including \emph{type safety} and a key modularity result, which we call  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e. in some  ``closed world'') are conserved in any further extended tycon context (i.e. in the ``open world''). Interestingly, type system providers need not necessarily provide  mechanized proofs to maintain these guarantees. Instead, the approach we take relies on type abstraction in the internal language. As a result, we are able to borrow the same results that underly modular reasoning in simply-typed languages like ML to reason modularly about typed language fragments. We describe  related work in Sec. \ref{prior-work} and conclude in Sec. \ref{discussion}.

\section{Overview of @$\lambda$}\label{atlam}\label{overview}
\paragraph{External Language}
Programmers interface with @$\lambda$ by writing \emph{external terms}, $e$. The syntax of external terms is shown in Figure \ref{syntax-EL}. %We will describe useful syntactic desugarings atop this syntax as we go on and review recent techniques that permit modularly introducing  desugarings like these in Sec. \ref{desugaring}. Our main focus here is on semantic extensions. 
The static and dynamic semantics are given simultaneously as a \emph{bidirectionally typed translation semantics}, i.e. the key judgements take the form:  \[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
\noindent

These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ and has  translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''. Note that our specifications in this paper are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$; these are not part of the judgement's syntax. In particular, note that the type is an ``output'' in the synthetic judgement, but an ``input'' in  the analytic judgement.

@@

We choose bidirectional typechecking, also sometimes called \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}, for two main reasons. The first is once again to justify the practicality of our approach: local type inference is increasingly being used in modern languages (e.g. Scala\todo{citation}) because it eliminates the need for type annotations in many situations while remaining decidable in more situations than whole-function type inference and providing what are widely perceived to be higher quality error messages\todo{citation from Dunfield}. Secondly, it will give us a clean way to reuse the generalized introductory form, $\eintro{\st}{\es}$, and its associated desugarings, at many types, in a manner that relates to recent mechanisms supporting type-specific syntax extensions \cite{TSLs}. For example, when we define regular string types, we will be able to reuse standard string literal syntax.

@@

The internal dynamics are specified also in a standard way as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}. The multi-step judgement $\iota \mapsto^{*} \iota\moutput$ is the reflexive, transitive closure of the stepping judgement and the evaluation judgement $\iota \Downarrow \iota'$ is defined iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$. Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity and key metatheoretic properties.

@@

\paragraph{Static Language}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-SL}
\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $m$ and $n$ over natural numbers.}\vspace{-5px}
\label{syntax-SL}
\end{figure}
The workhorse of @$\lambda$ is the \emph{static language}, which itself forms a typed lambda calculus where 
%External terms are classified by (external) \emph{types}. 
\emph{kinds}, $\kappa$, classify \emph{static terms}, $\sigma$.  The syntax of the SL is given in Figure \ref{syntax-SL}. The portion of the SL covered by the first row of kinds and first four rows of static terms in the syntax forms an entirely standard  functional programming language consisting of total functions, universal quantification over kinds, inductive kinds (constrained by the standard positivity condition to prevent non-termination), and products and sums. The reader can consider these as forming a total subset of ML or a simply-typed subset of Coq. The semantics also directly follow \cite{pfpl}, so we  omit the details here.  Only three new  kinds will be needed: $\kty$ (Sec. \ref{types}), $\kity$ (Sec. \ref{sec:type-translations}) and $\kitm$ (Sec. \ref{sec:introop}).

@@

The dynamic semantics of static terms is defined as a structural operational semantics by the stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, the value judgement $\sval{\st}{\argEnv}$ and the error judgement $\serr{\st}{\argEnv}$. Here, $\argEnv$ ranges over \emph{argument environments}, which we will  return to when considering opcons in Sec. \ref{external-terms}. The multi-step judgement $\smanystep{\st}{\argEnv}{\st\moutput}$ is the reflexive, transitive closure of the stepping judgement. The normalization judgement $\seval{\st}{\argEnv}{\st'}$ is defined iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$. %${\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code. We write $\st \Downarrow \st'$ iff $\snorm{\st}{\emptyset}{\cdot}{{\rightharpoonup}; \emptyset; \emptyset}{\st'}{\emptyset}{\cdot}$. \emph{Static values} are normalized static terms. Normalization can also raise an error (to indicate a type error in an external term, or a problem in a tycon definition, as we will discuss), indicated by the judgement $\serrX{\st}$. We omit error propagation rules.%We will refer to the relevant rules as we proceed. 
\begin{figure}[t]
\small
$\begin{array}{lrcl}
\textbf{tycons} & c & ::= & \rightharpoonup ~|~ \tc ~|~ \keyw{other}[m]\\
\textbf{tycon contexts} & \Phi & ::= & \cdot ~|~ \Phi, \tcdef{\tc}{\psi}{\theta}\\
\textbf{tycon structures} & \theta & ::= & \tcstruct{\st}{\omega} \\
\textbf{tycon sigs} & \psi & ::= & \tcsig{\kappa}{\chi}\\
\textbf{opcon structures} & \omega & ::= & \tcstructn{\st} ~|~ \tcstructc{\omega}{op}{\st}\\
\textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\theta}{op}{\kappa}\\
\end{array}$
\caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over user-defined tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
\label{syntax-TC}
\end{figure}
\begin{figure*}\begin{mathpar}
\small
\inferrule[k-parr]{
    \sofkX{\st}{\kprod{\kty}{\kty}}
}{
    \sofkX{\sty{\rightharpoonup}{\st}}{\kty}
}

@@

External types, or simply \emph{types}, are static values of kind $\kty$. The introductory form for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}. The syntax  for tycons in Figure \ref{syntax-TC} specifies that $c$ is either the built-in tycon governing partial functions, $\rightharpoonup$, a user-defined tycon name written in small caps, $\tc$, or an ``other'' tycon, $\keyw{other}[m]$ indexed by a natural number $m$ (to allow arbitrarily many such tycons). The kinding rules governing the form $\sty{c}{\st}$ are shown in Figure \ref{fig:types}. The dynamics are simple (see supplement): the index is eagerly normalized and errors propagate. We write $\istype{\st}{\Phi}$  iff $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\svalNA{\st}$.

@@

\paragraph{Examples}
As our first example, consider the user-defined tycon $\tcvar{rstr}$. The index kind of $\tcvar{rstr}$ is $\krx$, which classifies static regular expressions and is defined as an inductive sum kind in the usual way. Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which are statically known to be in the regular language specified by the type index  \cite{sanitation-psp14}. For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ classifies non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ classifies conference names.  The type indices are  here written using standard concrete syntax for concision. Previous work has shown how to define type-specific (here, kind-specific) syntax like this composably in libraries \cite{TSLs}. We define a tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.

@@

\subsection{Type Case Analysis}
A type $\st$ can be case analyzed against a known tycon $c$ using $\stycase{c}{\st}{\sx}{\st_1}{\st_2}$. If the value of $\st$ is constructed by $c$, its type index is bound to $\sx$ and the branch $\st_1$ is taken. For totality, a default branch, $\st_2$, must also be provided.  For example, the kinding rule for when $c$ is user-defined is below. 
\begin{mathpar}
\small
\inferrule[k-tycase]{
    \sofkX{\st}{\kty}\\
    \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
    \sofk{\kDelta}{\kGamma, \sx :: \ktyidx}{\Phi}{\st_1}{\kappa}\\
    \sofkX{\st_2}{\kappa}
}{
    \sofkX{\stycase{\tc}{\st}{\sx}{\st_1}{\st_2}}{\kappa}
}
\end{mathpar}
% \begin{mathpar}\small
% \inferrule[s-tycase-fail-2]{ }{
%     \sstep{\stycase{c}{\sotherty{m}{\tau}}{x}{\st_1}{\st_2}}{\argEnv}{\st_2}
% }
% \end{mathpar}

The rule for arrow types is analagous, but no rule for tycons of the form $\keyw{other}[m]$ is defined.

>>>>

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%extensible languages; module systems; type abstraction; typed compilation; type-level computation
\section{Introduction}\label{intro}
Typed programming languages are most often described as being composed of \emph{fragments}, each contributing to the language's concrete syntax, abstract syntax, static semantics and dynamic semantics. 
In his textbook, Harper organizes fragments around type constructors, describing each in a different chapter \cite{pfpl}. Languages are then identified by a set of type constructors, e.g. $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language that includes partial function types, polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed below).
Another related approach is to describe fragments with a simple calculus having a ``catch-all'' constant and base type to stand notionally for all other terms and types that may also be included in some future complete language.

@@

This is not an everyday problem for programmers only because fragments like those mentioned above are ``general purpose'': they make it possible to \emph{isomorphically embed}  many other fragments as ``libraries''. For example, list types need not be built in because they are isomorphic to the type $\iforall{\alpha}{\imu{t}{\iunit + (\alpha \times t))}}$ (datatypes in ML combine these into a single declaration construct). %Languages providing datatypes al\'a ML  are perhaps most directly oriented around  embeddings into this core. % Leaving the core language simple makes it easier to establish its metatheory and verify that tools, like compilers, are implemented correctly.% They also have encouraging connections to logic. %For example, we do not need to define the type constructor $\fvar{list}$ (indexed by a type) as a fragment (though it is possible to do so) because there is a user-defined datatype constructor, $\texttt{list}$, parameterized by a type, that is isomorphic. If we added the $\fvar{list}$ fragment to the language, it would be entirely redundant semantically: every well-typed term of a type constructed by $\fvar{list}$ corresponds to a well-typed term and a polymorphic recursive sum type such th%Although strings and numbers can be embedded as recursive sum types, it is recognized that this is impractical, so these are also usually included as primitives.
%Establishing an isomorphic embedding of a desirable fragment in terms of general-purpose fragments is not always possible, nor are such embeddings always practical.

Universality properties (i.e. ``Turing-completeness'') can be invoked to guarantee that an embedding that preserves a desirable fragment's dynamic semantics is possible, but an isomorphic embedding must also preserve the static semantics and, if defined, performance bounds specified using a cost semantics. This is not always possible. 
Embeddings are also sometimes too \emph{complex}, as measured by the cost of the extralinguistic computations that are needed to map in and out of the embedding and, if these  must be performed mentally by programmers, considering various human factors. %We will discuss specific examples below.
%When an embedding is too complex, abstraction providers have a few options, discussed in Sec. \ref{prior-work}. When an embedding is not possible, however, or when these options are  insufficient, 
%providers are often compelled to introduce these fragments by extending an existing language, thereby forming a new \emph{dialect}. To save effort, they may do so by forking existing artifacts or leverage tools like compiler generators or language frameworks, also reviewed in Sec. \ref{prior-work}. %Indeed, the proliferation of language dialects constructed for this reason might be taken as an evidence that the core language is not as ``general'', when considered comprehensively, as might be hoped. 
Each time a language is extended with a new fragment for one of these reasons, a new \emph{dialect} is born. Within the ML lineage, for example, dialects abound:
%Reynolds, in a remark that recalls the ``Turing tarpit'' of Perlis \cite{Perl82a}, summarizes the issue \cite{Reynolds94anintroduction}: 
%\begin{quote}
%To say that any reasonable function can be expressed by some program is not to say that it can be expressed by the most reasonable program. It is clear that the language requires a novel programming style. Moreover, it is likely that certain important functions cannot be expressed by their most efficient algorithms.
%\end{quote}
%

@@

This \emph{dialect-oriented} state of affairs is unsatisfying. %Language designers are burdened with needing to understand the complete metatheory of every fragment they add to their language to make sure it is not incompatible with existing fragments, so languages change rarely. This decreases the impact of many potentially useful innovations designed by fragment providers (e.g. many academic researchers) by making them unavailable to programmers. 
While programmers can choose from dialects supporting, e.g., a principled approach to distributed programming, or one that builds in support for statically reasoning about units of measure, one that supports both fragments may not be available. Using different dialects separately for different components of a program is untenable: components written in different dialects cannot always interface safely (i.e. a safe FFI, item 3 above, is needed).

These problems do not merit concern when a fragment is expressed as an isomorphic embedding (i.e. as a library) because modern \emph{module systems} enforce abstraction barriers that ensure that the isomorphism need only be established in the ``closed world'' of the module. For example, a module defining sets in ML can hold the representation of sets abstract, ensuring that any invariants maintained by the functions in the module (e.g. uniqueness, if using a list representation) will hold no matter which other modules are separately in use by a client.\todo{citation?} %Mechanisms that can help decrease the complexity of an embedding without violating abstraction barriers are thus valuable, and we will lead into our work in Sec. \ref{prior-work} by summarizing them.

@@

\paragraph{Contributions} In this paper, we take foundational steps towards this goal by  constructing a simple but surprisingly powerful core calculus, @$\lambda$ (the ``actively typed'' lambda calculus). %Despite its minimality, it can host a variety of practical semantic extensions like those described above, while maintaining strong metatheoretic guarantees and, crucially, providing modular reasoning properties. %a calculus introduced briefly in recent work on \emph{active type constructors}  \cite{atlang-gpce14}\todo{TR? Arxiv?}, reviewed in Sec. \ref{overview}. 
Its semantics are structured like those of many modern languages, consisting of an \emph{external language} (EL) governed by a {typed translation semantics} targeting a much simpler \emph{internal language} (IL). 
Rather than building in a monolithic set of external type constructors, however, the  semantics are indexed by a \emph{tycon context}. Each tycon defines the semantics of its operators via functions written in a \emph{static language} (SL) where types and translations are values. %, which provides a form of type-level computation. %The authors demonstrate the expressive power of this technique primarily with an implementation and a number of examples of fragments as libraries, and their core calculus .% (unlike systems that treat the type system as a ``bag of rules'', where non-determinism can arise).

We will begin by giving an overview of the organization  and main judgements of @$\lambda$ in Sec. \ref{atlam}, then discuss how types are constructed and introduce our two main examples, one defining labeled product types with a functional update operator, and the other regular string types, based on a recent core calculus style specification \cite{sanitation-psp14}, in Sec. \ref{types}. We describe how tycons control the semantics of their associated term-level operator constructors (opcons) in Sec. \ref{external-terms}.  We next give the key metatheoretic properties of the calculus in Sec. \ref{metatheory}, including \emph{type safety} and a key modularity result, which we call  \emph{conservativity}: any invariants that can be established about all values of a type under \emph{some} tycon context (i.e. in some  ``closed world'') are conserved in any further extended tycon context (i.e. in the ``open world''). Interestingly, type system providers need not necessarily provide mechanized proofs to maintain these guarantees. Instead, the approach we take relies on type abstraction in the internal language. As a result, we are able to borrow the same results that underly modular reasoning in simply-typed languages like ML to reason modularly about typed language fragments. We describe  related work in Sec. \ref{prior-work} and conclude with a brief discussion of expressiveness in Sec. \ref{discussion}.

\section{Overview of @$\lambda$}\label{atlam}\label{overview}
\paragraph{External Language}
Programmers interface with @$\lambda$ by writing \emph{external terms}, $e$. The abstract syntax of external terms is shown in Figure \ref{syntax-EL} (we will give various concrete desugarings in Sec. \ref{external-terms}). %We will describe useful syntactic desugarings atop this syntax as we go on and review recent techniques that permit modularly introducing  desugarings like these in Sec. \ref{desugaring}. Our main focus here is on semantic extensions. 
The static and dynamic semantics are given simultaneously as a \emph{bidirectionally typed translation semantics}, i.e. the key judgements take the form:  \[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
\noindent

These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ and has  translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''. Note that our specifications in this paper are intended to be algorithmic: we indicate ``outputs'' when introducing judgement forms by \emph{mode annotations}, $\moutput$; these are not part of the judgement's syntax. In particular, note that the type is an ``output'' only in the synthetic judgement.

@@

We choose bidirectional typechecking, also sometimes called \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}, for two main reasons. The first is once again to justify the practicality of our approach: local type inference is increasingly being used in modern languages (e.g. Scala\todo{citation}) because it eliminates the need for type annotations in many situations while remaining decidable in more situations than whole-function type inference and providing what are widely perceived to be higher quality error messages\todo{citation from Dunfield}. Secondly, it will give us a clean way to reuse the generalized introductory form, $\eintro{\st}{\es}$, and its associated desugarings, at many types. For example, regular string types will be able to use standard string literal syntax.

@@

The internal dynamics are specified as a structural operational semantics with a stepping judgement {\small $\iota \mapsto \iota\moutput$} and a value judgement {$\iota~\mathtt{val}$}. The multi-step judgement $\iota \mapsto^{*} \iota\moutput$ is the reflexive, transitive closure of the stepping judgement and the evaluation judgement $\iota \Downarrow \iota'$ is defined iff $\iota \mapsto^{*} \iota'$ and $\iota'~\mathtt{val}$. Both the static and dynamic semantics of the IL can be found in any standard textbook covering typed lambda calculi (we directly follow \cite{pfpl}), so we assume familiarity and key metatheoretic properties.

@@

\paragraph{Static Language}
\begin{figure}[t]
\small
\hspace{-5px}\input{syntax-table-SL}
\caption{Syntax of the static language (SL). Metavariable $\sx$ ranges over static term variables, $\kalpha$ and $\kvar$ over kind variables and $m$ and $n$ over natural numbers.}\vspace{-5px}
\label{syntax-SL}
\end{figure}
The workhorse of @$\lambda$ is the \emph{static language}, which itself forms a typed lambda calculus where 
%External terms are classified by (external) \emph{types}. 
\emph{kinds}, $\kappa$, classify \emph{static terms}, $\sigma$.  The syntax of the SL is given in Figure \ref{syntax-SL}. The portion of the SL covered by the first row of kinds and first four rows of static terms in the syntax forms an entirely standard  functional programming language consisting of total functions, universal quantification over kinds, inductive kinds (constrained by the standard positivity condition to prevent non-termination), and products and sums. The reader can consider these as forming a total subset of ML or a simply-typed subset of Coq. The semantics also directly follow \cite{pfpl}, so we  omit the details of this core here.  Only three new  kinds will be needed: $\kty$ (Sec. \ref{types}), $\kity$ (Sec. \ref{sec:type-translations}) and $\kitm$ (Sec. \ref{sec:introop}).

@@

The dynamic semantics of static terms is defined as a structural operational semantics by a stepping judgement $\sstep{\st}{\argEnv}{\st\moutput}$, a value judgement $\sval{\st}{\argEnv}$ and an error judgement $\serr{\st}{\argEnv}$. Here, $\argEnv$ ranges over \emph{argument environments}, which we will  return to when considering opcons in Sec. \ref{external-terms}. The multi-step judgement $\smanystep{\st}{\argEnv}{\st\moutput}$ is the reflexive, transitive closure of the stepping judgement. The normalization judgement $\seval{\st}{\argEnv}{\st'}$ is defined iff $\smanystep{\st}{\argEnv}{\st'}$ and $\sval{\st'}{\argEnv}$. %${\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code. We write $\st \Downarrow \st'$ iff $\snorm{\st}{\emptyset}{\cdot}{{\rightharpoonup}; \emptyset; \emptyset}{\st'}{\emptyset}{\cdot}$. \emph{Static values} are normalized static terms. Normalization can also raise an error (to indicate a type error in an external term, or a problem in a tycon definition, as we will discuss), indicated by the judgement $\serrX{\st}$. We omit error propagation rules.%We will refer to the relevant rules as we proceed. 
\begin{figure}[t]
\small
$\begin{array}{lrcl}
\textbf{tycons} & c & ::= & \rightharpoonup ~|~ \tc ~|~ \keyw{other}[m]\\
\textbf{tycon contexts} & \Phi & ::= & \cdot ~|~ \Phi, \tcdef{\tc}{\psi}{\theta}\\
\textbf{tycon structures} & \theta & ::= & \tcstruct{\st}{\omega} \\
\textbf{tycon sigs} & \psi & ::= & \tcsig{\kappa}{\chi}\\
\textbf{opcon structures} & \omega & ::= & \tcstructn{\st} ~|~ \tcstructc{\omega}{op}{\st}\\
\textbf{opcon sigs}& \chi & ::= & \introsig{\kappa} ~|~ \opsigS{\theta}{op}{\kappa}\\
\end{array}$
\caption{Syntax of tycons. Metavariables $\tc$ and $\opname{op}$ range over user-defined tycon and opcon names, respectively, and $m$ ranges over natural numbers.}
\label{syntax-TC}
\end{figure}
\begin{figure*}\begin{mathpar}
\small
\inferrule[k-parr]{
    \sofkX{\st}{\kprod{\kty}{\kty}}
}{
    \sofkX{\sty{\rightharpoonup}{\st}}{\kty}
}

@@

External types, or simply \emph{types}, are static values of kind $\kty$. The introductory form for kind $\kty$ is $\sty{c}{\st}$, where $c$ is a \emph{tycon} and $\st$ is the \emph{type index}. The syntax  for tycons given in Figure \ref{syntax-TC} specifies that $c$ is either the built-in tycon governing partial functions, $\rightharpoonup$, a user-defined tycon name written in small caps, $\tc$, or an ``other'' tycon, $\keyw{other}[m]$ indexed by a natural number $m$ (to allow arbitrarily many such tycons). The kinding rules governing the form $\sty{c}{\st}$ are shown in Figure \ref{fig:types}. The dynamics are simple (see supplement): the index is eagerly normalized and errors propagate. We write $\istype{\st}{\Phi}$  iff $\sofkn{\emptyset}{\emptyset}{\Phi}{0}{\st}{\kty}$ and $\svalNA{\st}$.

@@

\paragraph{Examples}
As our first example, consider the user-defined tycon $\tcvar{rstr}$. The index kind of $\tcvar{rstr}$ is $\krx$, which classifies static regular expression patterns and we assume is defined as an inductive sum kind in the usual way. Types constructed by $\tcvar{rstr}$ will classify \emph{regular strings}, which are statically known to be in the regular language specified by the type index  \cite{sanitation-psp14}. For example, $\stx{title} := \sty{\tcvar{rstr}}{\concrx{.+}}$ will classify non-empty strings and $\stx{conf} := \sty{\tcvar{rstr}}{\concrx{[A-Z]+ \digit\digit\digit\digit}}$ will classify conference names.  The type indices are  here written using standard concrete syntax for concision. Previous work has shown how to define type-specific (here, kind-specific) syntax like this composably in libraries \cite{TSLs}. We define a tycon context containing only the definition of $\tcvar{rstr}$, $\Phi_\text{rstr} := \tcdef{\tcvar{rstr}}{\tcsig{\krx}{\chi_\text{rstr}}}{\theta_\text{rstr}}$.

@@

\subsection{Static Type Case Analysis}
A type $\st$ can be case analyzed using $\stycase{c}{\st}{\sx}{\st_1}{\st_2}$. If the value of $\st$ is constructed by $c$, its type index is bound to $\sx$ and the branch $\st_1$ is taken. For totality, a default branch, $\st_2$, must also be provided.  For example, the kinding rule for when $c$ is user-defined is below. 
\begin{mathpar}
\small
\inferrule[k-tycase]{
    \sofkX{\st}{\kty}\\
    \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\theta} \in \Phi\\
    \sofk{\kDelta}{\kGamma, \sx :: \ktyidx}{\Phi}{\st_1}{\kappa}\\
    \sofkX{\st_2}{\kappa}
}{
    \sofkX{\stycase{\tc}{\st}{\sx}{\st_1}{\st_2}}{\kappa}
}
\end{mathpar}
% \begin{mathpar}\small
% \inferrule[s-tycase-fail-2]{ }{
%     \sstep{\stycase{c}{\sotherty{m}{\tau}}{x}{\st_1}{\st_2}}{\argEnv}{\st_2}
% }
% \end{mathpar}
The rule for $c{=}{\rightharpoonup}$ is analagous, but no rule for $c{=}\keyw{other}[m]$ is defined (these always take the default branch).

Put another way, types can be thought of as arising from a distinguished ``open datatype'' defined by the tycon context\todo{citation}. %The dynamics (see supplement) are again straightforward, consistent with this intuition.% We will see an example of its use in Sec. \ref{sec:targops}.
\begin{figure}\hfill \fbox{$\vdash \Phi$}\vspace{-25px}\begin{mathpar}
\small
\inferrule[tcc-ext]{
    \vdash \Phi\\
    \tc \notin \text{dom}(\Phi)\\
    \keq{\emptyset}{\ktyidx}\\
    \sofkz{\emptyset}{\emptyset}{\Phi}{\stx{schema}}{\ktyidx \rightarrow \kity}\\\\
    \vdash_{\Phi,  \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}} \omega \sim \tcsig{\ktyidx}{\chi}
}{
    \vdash \Phi, \tcdef{\tc}{\tcsig{\ktyidx}{\chi}}{\tcstruct{\stx{schema}}{\omega}}
}\vspace{-8px}
\end{mathpar}
\caption{Tycon context well-definedness. We omit the trivial case for when $\Phi=\cdot$ for concision.}
\label{fig:tycon-ctxs}
\vspace{-5px}
\end{figure}
\subsection{Tycon Context Well-Definedness}
The tycon context well-definedness judgement, $\vdash \Phi$, shown in Figure \ref{fig:tycon-ctxs}, requires that all tycon names are unique and performs additional checks,  described below.

\subsection{Type Equivalence}\label{sec:type-equivalence}
The first check simplifies the handling of type equivalence: type index kinds must be \emph{equality kinds}, i.e. those for which semantic equivalence implies syntactic equality at normal form. We define these by the judgement $\keq{\kDelta}{\kappa}$ (see supplement). 
Equality kinds are similar to equality types as found in Standard ML\todo{citation}. The main implication of this choice is that type indices cannot contain static functions.
